---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions,echo=TRUE}
knitr::opts_chunk$set(echo=TRUE)
```

```{r computetime,results='hide'}
Sys.setlocale(category = "LC_ALL", locale = "en_US.utf8")
last_compute_time<-format(Sys.time(),"%a %b %d %Y %X ")
data_archive_filename<-"activity.zip"
data_filename<-"activity.csv"
```
last computed at : `r last_compute_time` 

## Loading and preprocessing the data
I will first extract data archive  `r data_archive_filename` in current directory
```{r}
unzip(data_archive_filename)
```
an load the data inside a dataframe variable called df
```{r}
df<-read.csv(data_filename)
```
here is a summary of the original data
```{r summary}
library(knitr)
kable(summary(df))
```


## What is mean total number of steps taken per day?
```{r steps_per_day}
library(dplyr)

df_byday<-df %>%
  dplyr::group_by(date) %>%
  summarise( num_steps_value = n(),
             total_steps_per_day = sum(steps)
            )
```
here is an histogram of the total number of steps per day
```{r histogram_steps_per_day}
hist(df_byday$total_steps_per_day)
```
and here is the mean total number of steps taken per day
```{r mean_steps_per_day}
mean(df_byday$total_steps_per_day,na.rm = TRUE)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
